Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-15 paper-1/manuscript.tex
Editing

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{}
    \label{fig2}
\end{figure*}
\textbf{Fig. 2} Deviance detection in vitro and in silico. \textbf{a} Response of a 2 \times 4~mm neuronal cell culture on a multi-electrode array, adapted from \cite{Kubota2021-dx}. Curves represent the average number of spikes per electrode in each 10~ms bin in response to target stimulation in the indicated sequence. Unlike in our study, oddball sequences here were split 10\%/90\%, and 10 equally frequent stimuli were used for the msc sequence. \textbf{b} Sample model response, showing average network-wide spike counts after stimulation with target (A) in the indicated sequence (same color code as panel \textbf{a}). Both in vitro and in the model, the dev response is very large, followed by an intermediate msc response and a low std response. \textbf{c} Mismatch, SSA and deviance detection indices (see Equation \ref{eqn-ddi}) across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the 95\% confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points

@@

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than other responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences.
To show this in the sample network, we correlated the number of spikes in dev target trials with the spike-triggered depression variable $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was very clear in the trial immediately following target (Pearson's \rho = -0.984, p = 0), it was almost completely lost by the next trial (\rho = -0.0525, p = 0.138) under the influence of both decay and unrelated activity in the intervening trial. In fact, even without intervening activity, inputs to $x_j$ would decay to a negligible $e^\frac{1 s}{\tau_x} = 0.0013$ within two trials.
Thus, STD could not have any stimulus-specific effects across dev target trials or msc trials.

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{}
    \label{fig2}
\end{figure*}
\textbf{Fig. 2} Deviance detection in vitro and in silico. \textbf{a} Response of a 2 $\times$ 4~mm neuronal cell culture on a multi-electrode array, adapted from \cite{Kubota2021-dx}. Curves represent the average number of spikes per electrode in each 10~ms bin in response to target stimulation in the indicated sequence. Unlike in our study, oddball sequences here were split 10\%/90\%, and 10 equally frequent stimuli were used for the msc sequence. \textbf{b} Sample model response, showing average network-wide spike counts after stimulation with target (A) in the indicated sequence (same color code as panel \textbf{a}). Both in vitro and in the model, the dev response is very large, followed by an intermediate msc response and a low std response. \textbf{c} Mismatch, SSA and deviance detection indices (see Equation \ref{eqn-ddi}) across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the 95\% confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points

@@

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than other responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences.
To show this in the sample network, we correlated the number of spikes in dev target trials with the spike-triggered depression variable $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was very clear in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the influence of both decay and unrelated activity in the intervening trial. In fact, even without intervening activity, inputs to $x_j$ would decay to a negligible $e^\frac{1 s}{\tau_x} = 0.0013$ within two trials.
Thus, STD could not have any stimulus-specific effects across dev target trials or msc trials.
